Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-02-24 paper/manuscript.Rmd
comment out new fig
fix typos

The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
observed an unmarked adult male caribou swim between Western and Eastern Indian
Islands (Figure 1), a swim of at least 470 m which took ~9 minutes (~52 m per
minute). This observation prompted us to investigate whether swimming behaviour
was common among GPS radio-collared caribou in the population (n = 29 adult
females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario, Canada). We
identified swimming events as two consecutive GPS (2-hour relocation rates)
locations from an individual occurred on different islands. Northeastern
Newfoundland typically experiences pack ice during winter and caribou are known
locally to travel between islands by walking over the ice. We restricted our GPS
data to the ice-free period of the year (April 1 - November 25) <!--confirm
dates --> and we identified three adult female caribou that collectively had 86
swimming events (Figure 2). In addition to our own visual and remotely sensed
observations, residents of Fogo Island have also reported observing caribou
swimming between islands on numerous occaisions.

<!-- add numbers in here -->
In total, these 86 swimming events occured in month (n = Y), month (n = Y), and
month (n = Y), while the remaining events were in month, month, and month (n =
Y). On average, caribou swam between islands approximately ever 24 days, but
some individuals swam up to XX times per week/month/whatever. <!--Is it better
to just state our results here, or give this sort of teaser? Initially I had "86
events by 3 individuals" here, but then we ignore that for a couple paragraphs
of context before getting back into it... which seemed weird-->

@@

In total, 3 of 29 collared female caribou swam between islands in the Fogo
archipelago. These swimming events were relatively frequent, on average occuring
once every 24 <!--confirm--> days during ice-free seasons. However, despite
relatively frequent swimming events for these three individuals, we were unable
to identify any swimming events for the other 26 collared females. It remains
unclear why ocean swimming is relatively common among ~10% of our marked
population, while the other 90% never swam. Of the three collared individuals,
two were collared on Western Indian Island and the third was collared on Fogo
Island. It is possible that individuals born on, or that live on, Eastern and
Western Indian Island swim are more likely to swim between islands because
inter-island distances are shorter than between Western Indian and Fogo Islands
(Figure 2). Furthermore, it is unclear whether swimming is more common among
males or females. Our initial observation  was of an adult male, but our GPS
data only includes females. <!-- QW: not sure how to finish this off -->
<!-- QW: I'm not sure we need to say this.. 
It is theoretically possible that these other
females moved rapidly back and forth between islands within the 2 hour fix rate
of our GPS collars, which we would be unable to detect. We consider this highly
unlikely given the range of residency times for caribou on any given island (min
- max: XX - XX). <!--double check-->
<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->

@@

![**Figure 1**: Image of swimming unmarled adult male caribou (*Rangifer tarandus*) from Western to Eastern Indian Island taken on 30 May 2017.](../graphics/IMG_1562.JPG)

@@

![**Figure 2**: Map of the Fogo Island archipelago. Fogo Island is the largest
island in the archipelago (237.71 km^2^), but there are at least three other
large islands: Western Indian (77.6 km^2^), Eastern Indian (38.7 km^2^) and
Change (XX km^2^) Islands, as well as numerous smaller islands, including
Blundon’s (1.18 km^2^), North Long (1.01 km^2^), South Long (0.48 km^2^), Kate’s
(1.64 km^2^), and Brother’s (1.59 km^2^) Islands. Habitats are similar across
the archipelago, consisting largely of coniferous and mixed forests of balsam
fir (*Abies balsamea*), black spruce (*Picea mariana*), and white birch 
(*Betula papyrifera*) as well as bogs, lakes, lichen and rocky
barrens.]()<!--(../graphics/IMG_1562.JPG)-->

>>>>

The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
observed an unmarked adult male caribou swim between Western and Eastern Indian
Islands (Figure 1), a swim of at least 470 m which took ~9 minutes (~52 m per
minute). This observation prompted us to investigate whether swimming behaviour
was common among GPS radio-collared caribou in the population (n = 29 adult
females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario, Canada). We
identified swimming events as two consecutive GPS (2-hour relocation rates)
locations from an individual occurred on different islands. Northeastern
Newfoundland typically experiences pack ice during winter and caribou are known
locally to travel between islands by walking over the ice. We restricted our GPS
data to the ice-free period of the year (April 1 - November 25) <!--confirm
dates --> and we identified three adult female caribou that collectively had 86
swimming events (Figure 2). In addition to our own visual and remotely sensed
observations, residents of Fogo Island have also reported observing caribou
swimming between islands on numerous occasions.

<!-- add numbers in here -->
In total, these 86 swimming events occurred in month (n = Y), month (n = Y), and
month (n = Y), while the remaining events were in month, month, and month (n =
Y). On average, caribou swam between islands approximately ever 24 days, but
some individuals swam up to XX times per week/month/whatever. <!--Is it better
to just state our results here, or give this sort of teaser? Initially I had "86
events by 3 individuals" here, but then we ignore that for a couple paragraphs
of context before getting back into it... which seemed weird-->

@@

In total, 3 of 29 collared female caribou swam between islands in the Fogo
archipelago. These swimming events were relatively frequent, on average occurring
once every 24 <!--confirm--> days during ice-free seasons. However, despite
relatively frequent swimming events for these three individuals, we were unable
to identify any swimming events for the other 26 collared females. It remains
unclear why ocean swimming is relatively common among ~10% of our marked
population, while the other 90% never swam. Of the three collared individuals,
two were collared on Western Indian Island and the third was collared on Fogo
Island. It is possible that individuals born on, or that live on, Eastern and
Western Indian Island swim are more likely to swim between islands because
inter-island distances are shorter than between Western Indian and Fogo Islands
(Figure 2). Furthermore, it is unclear whether swimming is more common among
males or females. Our initial observation  was of an adult male, but our GPS
data only includes females. <!-- QW: not sure how to finish this off -->
<!-- QW: I'm not sure we need to say this.. 
It is theoretically possible that these other
females moved rapidly back and forth between islands within the 2 hour fix rate
of our GPS collars, which we would be unable to detect. We consider this highly
unlikely given the range of residency times for caribou on any given island (min
- max: XX - XX). <!--double check-->
<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->

@@

![**Figure 1**: Image of swimming unmarked adult male caribou (*Rangifer tarandus*) from Western to Eastern Indian Island taken on 30 May 2017.](../graphics/IMG_1562.JPG)

@@

<!--![**Figure 2**: Map of the Fogo Island archipelago. Fogo Island is the largest
island in the archipelago (237.71 km^2^), but there are at least three other
large islands: Western Indian (77.6 km^2^), Eastern Indian (38.7 km^2^) and
Change (XX km^2^) Islands, as well as numerous smaller islands, including
Blundon’s (1.18 km^2^), North Long (1.01 km^2^), South Long (0.48 km^2^), Kate’s
(1.64 km^2^), and Brother’s (1.59 km^2^) Islands. Habitats are similar across
the archipelago, consisting largely of coniferous and mixed forests of balsam
fir (*Abies balsamea*), black spruce (*Picea mariana*), and white birch 
(*Betula papyrifera*) as well as bogs, lakes, lichen and rocky
barrens.]()<!--(../graphics/IMG_1562.JPG)-->
